---
title: "Getting started with R"
author: "Ethan Weed"
output: 
  html_document:
    df_print: default
    number_sections: no
    self_contained: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<!-- From https://stackoverflow.com/a/37839683/1297830 -->
<link rel="stylesheet" type="text/css" href="hideOutput.css">
<script src="hideOutput.js"></script>

```{r echo=FALSE}
# Options for building this document
knitr::opts_chunk$set(
  fig.height = 4,
  fig.width = 6,
  fig.align = 'center',
  message = FALSE,
  warning = FALSE
)
```

# RStudio
The R programming language is an important tool for exploring, visualizing, and analyzing data. It is much more powerful and flexible than point-and-click statistical analysis programs or spreadsheets, but it does take some getting used to, especially if you are unfamiliar with programming. I reccomend RSstudio for working with R. RStudio is a so-called IDE (integrated development environment) for the R language. An IDE is a program that makes it easier to write code and get your work done. RStudio has many many features that make life easier for us when working with R. You can even use RStudio to make slides and webpages. The words you are reading now were typed into RStudio.

# Variables
Variables are fundamental to programming in any language. A variable can be thought of as a container, or as a label. Here is an example:

```{r}
a <- "Hello world!"
```

R uses the syntax ``` <- ``` to show that we are putting something into a variable. Here we are putting the words "Hello world!" into the variable ```a```. Once a variable has been defined, we can do different things with it:

```{r}
a <- "Hello world!"
b <- 3
c <- 2

print(a)
b + c
b * c
b / c
```

## Variable types

There are different kinds of variables, and they can do different things:

```{r}
a <- "3.14159"
b <- 3.14159

class(a)
class(b)

```

In this case, ``a`` is a **character**, while ``b`` is a **number**. We could use ``b`` to do math, but we couldn't do math with `à``. 

### Dataframes

One of the most useful variable types for data analysis in R is the dataframe. A dataframe is a little bit like a spreadsheet, in that it has rows and columns, but we interact with is a ittle differently than we do with a spreadsheet like Excel. In Excel, you can just go in and click on a cell and change whatever you want. You can add plots on top of your data, and mess around however you please. this can be great for taking a quick look at data, but it is no good for actual analysis, becasue there is too much room for human error. It is too easy to accidently change something and mess up all your data. Dataframes in R are a little more structured than a spreadsheet, and because we interact with them *programatically*, by writing code, even if we mess something up, we can always trace back our steps, and figure out what went wrong.

The easiest way to get your data into R is by importing it from a .csv file. Click the button to show the entire dataframe. Here we look at data from students carrying out a version of the Stroop color-naming task.

<div class="fold o">
```{r}
df <- read.csv("/Users/ethan/Documents/GitHub/ethanweed.github.io/r-tutorials/data/Stroop-raw-over-the-years.csv")
df
```
</div>


There are 104 rows of data, and maybe we only want to work with the data from 2015. We can easily make a new dataframe, with only the 2015 data:
<div class="fold o">
```{r}
df_2015 <- subset(df, Year == 2015)
df_2015
```
</div>

You can use the ```$``` symbol to select a single column from the dataframe, using the column header:


```{r}
Reading_no_interference <- df_2015$Reading_NoInt
Reading_no_interference
```


# Exploring data with ggplot

One of the best ways to understand our data is to plot it. R has lots of tools for plotting data, and the package *ggplot* is one of the best. ""GG" stands for "Grammar of Graphics". This way of building plots makes it easy to mix and match elements, so you and make just the plot you want to show your data. Ggplot can be installed as a separate package, but it is also part of a package called *tidyverse* which has lots of great tools for wrangling data. Tidyverse (and other packages) can be installed through RStudio's package manager ("packages" tab in the lower right frame of the RStudio window) or by typing ```install.packages("tidyverse")``` 


```{r}
# Load packages for data handling and plotting
library(tidyverse)

# clear global environment
rm(list = ls())

df <- read.csv("/Users/ethan/Documents/GitHub/ethanweed.github.io/r-tutorials/data/Stroop-raw-over-the-years.csv")
```


The function ``head`` lets us see just the top of the dataframe. This is handy for just getting a quick peek at the structure of the data, without having to display all of it:


```{r}
head(df)
```


These particular data are stored in the so-called "wide" format, in which each row represents a single participant, and each column shows a different kind of information or measurement about the participant. To explore these data, it will be useful to transform them to the "long" format. We can easily do this using the function ``gather`` from the "tidyverse" package:

```{r}
# make a new dataframe with the same data in the "long" format.
df_long <- gather(df, key = "Condition", value = "Time", -Year)
head(df_long)
```

If you want, you can check that these are the same data, just arranged in a different way:

<div class="fold o">
```{r}
# display all the data in the original wide format
df

# display all the data in long format
df_long
```
</div>

## Bar plots
There are many different ways these data could be represented graphically. A common way to look at data from different experimental conditions is to make a bar plot.
```{r}
ggplot(df_long, aes(x = Condition, y = Time)) +
  geom_bar(stat = "identity")
```

Maybe we would like to see the columns as horizontal bars (with coordinates flipped):
```{r}
ggplot(df_long, aes(x = Condition, y = Time)) +
  geom_bar(stat = "identity") +
  coord_flip()
```

## Scatter plots
One problem with bar plots is that they usually only represent the mean value, and give little feeling for what the underlying data look like. With a scatter plot, we can see each of our individual data points. While we're at it, let's add some color just to make it pretty:
```{r}
ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_point()
```

## Box plots
Box plots are nice for adding some descriptive statistics to the plot. Here, the horizontal line shows the median value, and the top and bottom parts of the box (the hinges) show the 2nd and 3rd quartiles. The vertical lines (whiskers) give an indication of outliers (for more details, type ``? geom_boxplot`` in the console): 
```{r}
ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_boxplot()
```

We can also layer different types of plots on top of each other, e.g. we can plot all the poins on top of our boxplots:
```{r}
ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_boxplot() +
  geom_point()
```

## Jitter
There are a lot of points lying on top of each other, so it is hard to see them all. One way to see them is to add "jitter" to spread them out a bit:
```{r}
ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_boxplot() +
  geom_jitter()
```

## Violin plots
Another method is to show probability density using violin plots:
```{r}
ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_boxplot() +
  geom_violin()
```

If we flip them on their sides, they kinda look like X-wing fighters, or maybe geese. Did we just invent the goose plot?
```{r}
ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_boxplot() +
  geom_violin() +
  coord_flip() +
  theme_classic()
```

## Combined plots
We could also combine several of our plots into one plot, e.g. to make it easier to compare them. Click the button to see the code. Although it looks like a lot, you will see it is mostly just copy/paste.

<div class="fold s">
```{r}
# load the library "patchwork". We will use this to combine the plots. Another option is the function "ggarrange" from the package "ggpubr". Both are good.
library(patchwork)

# theme_classic takes away the background grid for a cleaner look. There are many other themes. Try them, and find one you like. There are also many different color options.

# theme(legend.position="none") takes away the legends, which in this case provided redundant information, and cluttered up the plot. There are many other theme options to make your plot look just the way you want.

# make a barplot and call it "bar". Notice we have to write "fill" instead of "color", otherwise it will look bad. Try and see.
bar <- ggplot(df_long, aes(x = Condition, y = Time, fill = Condition)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  coord_flip() +
  theme(legend.position="none")

# make a boxplot and call it "box"  
box <- ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_boxplot() +
  coord_flip() +
  theme_classic() +
  theme(legend.position="none")

# make a scatterplot and call it "scatter"
scatter <- ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_point() +
  coord_flip() +
  theme_classic() +
  theme(legend.position="none")

# make a "gooseplot" and call it "goose". Just to be clear, there is no such thing as a gooseplot. I just made that up.

goose <- ggplot(df_long, aes(x = Condition, y = Time, color = Condition)) +
  geom_boxplot() +
  geom_violin() +
  coord_flip() +
  theme_classic() +
  theme(legend.position="none")

# Because we have "patchwork" loaded, we can just "add" the plots together.
bar + box + scatter + goose



```
</div>

# Correlations

One of the most basic things a researcher is interested in is seeing whether there is any relationship between variables. 

<div class="fold s">
```{r}
rm(list = ls())

library(tidyverse)
library(patchwork)

df <- read.csv("/Users/ethan/Documents/GitHub/ethanweed.github.io/r-tutorials/data/nettle_1998.csv")
df <- rename(df, "Number_of_Languages" = "Langs")

p1 <- ggplot(df, aes(Area, Population)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  geom_text(aes(label = Country)) +
  theme_classic()

p2 <- ggplot(df, aes(Number_of_Languages, Population)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic()

p1

```
</div>






